Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-16 write-ups/PLDI/paper.tex
Crunch.
updated paper graphs to use N instead of M and added all other graphs

The number of \emph{active workgroups}---workgroups executing the
kernel---is denoted $M$.  Active workgroups
have consecutive ids in the range $[0, M-1]$.
Initially, at least one workgroup is active;
if necessary the scheduler must postpone the kernel until
some compute unit becomes available.

@@

Because workgroups of a cooperative kernel are fairly scheduled, so a
global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
and $\resizingglobalbarrier$.

@@

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\end{figure}

@@

For ease of implementation, our prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
We
assume all variables declared at the root lexical scope of a
cooperative kernel are implicitly $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
of the kernel.  With these restrictions we can experiment
with interesting irregular algorithms (see
\mysec\ref{sec:experiments}).  A non-mock implementation of
coopeartive kernels would not use the megakernel approach, so we do not deem
the engineering effort associated with lifting these restrictions to
be worthwhile.

@@

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. For multi-kernel
kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
primitives via a per kernel \emph{context}, which records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to other primitives.  It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation.

@@

This barrier implementation is suboptimal becuase workgroups only
execute $\offerkill$ once per barrier call and, depending on
order of arrival, it is possible
that only one workgroup is killed per barrier call, preventing the scheduler from gathering workgroups quickly.

@@

\begin{table}[t]
\small
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\end{table}

\begin{table}[h]
\footnotesize
\centering
\begin{tabular}{ l l l l }
Chip & HD520 & HD5500 & Iris\\
\hline
\# CU  & 24 & 24 & 47 \\
Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
Host CPU & i3-6100U & i7-5600U & i3-5157U \\
\end{tabular}
\caption{The Intel GPU platforms used in our evaluation}
\label{tab:chipstested}
\end{table}

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
indicate how many of the original number of barriers are changed to
resizing barriers, and how many variables need to be transmitted.  In
all cases all barriers were converted, except for bc which contains
barriers deeper in the call stack, a case not supported by our
prototype although these barriers could in principle be
converted. The remaining 2 algorithms are workstealing applications:
each required the addition of $\offerfork$ and $\offerkill$ at the start
of the main loop, and no variables needed to be transmitted.  For all
these examples the porting to cooperative kernels was similarly
straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines} for our running examples. Most graph come with 2
different data sets as input, leading to 13 application/input pairs in
total.

@@

@@

\paragraph{Results}
For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and inputs, with averages and
maxima computed over 10 runs per benchmark. For the maximum slowdowns,
we indicate which application and input was responsible. The slowdown is
below 2 even in the worst case, and closer to 1 on average. We consider
these results as encouraging, especially since the performance of our
prototype could clearly be improved upon in a native implementation.

@@

Co-scheduling non-cooperative kernels that compute their workload using a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
of any of our three workloads. For example, a heavy workload averages a
period of 939 ms. As more workgroups are dedicated to non-cooperative
kernels, they execute quickly enough to be scheduled at the expected
period. More concretely, for the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to meet
their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a mean period of 88 ms). Finally, if half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time overhead of the cooperative kernel.

@@

\ADComment{Tyler: show all the graphs.}

>>>>

The number of \emph{active workgroups}---workgroups executing the
kernel---is denoted $M$.  The active workgroups
have consecutive workgroup ids in the range $[0, M-1]$.
Initially, at least one workgroup is active;
if necessary the scheduler must postpone execution of the kernel until
a some compute unit becomes available.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, as illustrated in the graph traversal example of \mysec\ref{sec:openclexamples} and \myfig\ref{fig:graphsearch}.
Because a global barrier requires fair scheduling of workgroups it cannot be safely implemented in OpenCL, CUDA or HSA, and thus is not provided as a primitive in any of these programming models.

Workgroups of a \emph{cooperative} kernel \emph{are} fairly scheduled, so a
global barrier primitive can be provided, and we specify two variants: $\globalbarrier$
and $\resizingglobalbarrier$.

@@

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing global barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\end{figure}

@@

For ease of implementation, our current prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
Rather than adding language support for $\transmit$ annotations, we
assume that all variables declared at the root lexical scope of a
cooperative kernel are $\transmit$-annotated; this allows our
implementation to piggy-back on top of Clang (see below) without
requiring front-end changes.  Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
of the kernel.  With these restrictions we can experiment
with interesting irregular algorithms (see
\mysec\ref{sec:experiments}).  A non-mock implementation of
coopeartive kernels would not use the megakernel approach, so we do not deem
the engineering effort associated with lifting these restrictions to
be worthwhile.

@@

When a workgroup executes a kernel (cooperative or otherwise), it must
be able to query a consistent context to obtain a workgroup id value
and the
number of workgroups executing the kernel. For multi-kernel
kernel execution, we emulate the $\getnumgroups$ and $\getgroupid$ 
primitives via a per kernel \emph{context}.  For a given kernel this records: the number of active groups $M$ (irrelevant for a non-cooperative kernel), the total number of requested groups $N$, and a mapping from the native workgroup id to a new workgroup
id valid for the kernel.

@@

In \mysec\ref{sec:semantics}, we defined the semantics of a resizing barrier in terms
of a series of calls to $\globalbarrier$, $\offerfork$ and $\offerkill$.
 It is possible, however to implement the resizing with
only one call to a global barrier with $\offerfork$ and $\offerkill$
baked into the implementation.

@@

This barrier implementation is suboptimal becuase workgroups only
execute $\offerkill$ once per global barrier call and, depending on
the order in which workgroups arrive at the barrier, it is possible
that only one workgroup is killed per barrier call.  This may not
allow the scheduler to gather workgroups required to execute another
task in a timely fashion.

@@

\begin{table}
\small
\centering
\begin{tabular}{ l r r r r r r}
App. & barriers & kill & fork & transmit & LoC & inputs\\
\hline
\rowcolor{Gray1}
color & 2 / 2 & 0 & 0 & 4 & 55 & 2\\
\rowcolor{Gray1}
mis & 3 / 3 & 0 & 0 & 0 & 71 & 2\\
\rowcolor{Gray1}
bc & 3 / 6 & 0 & 0 & 3 & 150 & 2\\
\rowcolor{Gray1}
p-sssp & 3 / 3 & 0 & 0 & 0  & 42 & 1\\
\rowcolor{Gray2}
bfs & 2 / 2 & 0 & 0  & 4  & 185 & 2\\
\rowcolor{Gray2}
l-sssp & 2 / 2 & 0 & 0  & 4  & 196 & 2\\
\rowcolor{Gray3}
octree & 0 / 0 & 1 & 1 & 0 & 213 & 1 \\
\rowcolor{Gray3}
game & 0 / 0 & 1 & 1 & 0 & 308 & 1 \\
\end{tabular} \\
\vspace{.2cm}
\crule[Gray1]{.2cm}{.2cm} Pannotia \hspace{.4cm} \crule[Gray2]{.2cm}{.2cm} Lonestar GPU  \hspace{.4cm}  \crule[Gray3]{.2cm}{.2cm} work stealing
\caption{Blocking GPU applications investigated}
\label{tab:applications}
\end{table}

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.  We
indicate how many of the original number of barriers are changed to
resizing barriers, and how many variables need to be transmitted.  In
all cases all barriers were converted, except for bc which contains
barriers deeper in the call stack, a case not supported by our
prototype. Still, these barriers could in principle also be
converted. The remaining 2 algorithms are workstealing applications:
each required the addition of $\offerfork$ and $\offerkill$ at the start
of the main loop, and no variables needed to be transmitted.  For all
these examples the porting to cooperative kernels was similarly
straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines} for our running examples, and did
not require reshaping of the original algorithm. Most graph come with 2
different data sets as input, leading to 13 application/input pairs in
total.

@@

\begin{table}
\footnotesize
\centering
\begin{tabular}{ l l l l }
Chip & HD520 & HD5500 & Iris\\
\hline
\# CU  & 24 & 24 & 47 \\
Driver   & 20.19.15.4501 & 10.18.15.4281 & 20.19.15.4463 \\
Host CPU & i3-6100U & i7-5600U & i3-5157U \\
\end{tabular}
\caption{The Intel GPU platforms used in our evaluation}
\label{tab:chipstested}
\end{table}

@@

\paragraph{Results}
For each GPU, \mytab\ref{tab:overhead} shows the geometric mean and
maximum slowdown across all applications and inputs, with averages and
maxima computed over 10 runs per benchmark. For the maximum slowdowns,
we indicate which application and input was responsible. The slowdown is
below 2 even in the worst case, and closer to 1 on average. We consider
these results as encouraging, especially since the performance of our
prototype could clearly be improved upon if cooperative scheduling was
implemented directly in the GPU driver.

@@

Co-scheduling non-cooperative kernels that compute their workload using a single workgroup leads to almost no overhead, but the period is far too high to meet the needs
of any of our three workloads. For example, a heavy workload averages a
period of 939 ms. As more workgroups are dedicated to non-cooperative
kernels, they execute quickly enough to be scheduled at the expected
period. More concretely, for the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to meet
their goal period (70 and 40 ms resp.). However, a quarter of the
workgroups are not sufficient to meet the goal for the heavy workload
(giving a mean period of 88 ms). Finally, if half of the workgroups are
allocated to the non-cooperative kernel, the heavy workload gets
averages \~10\% over its goal period (mean of 44 ms). If asking for all-but-one, the heavy workload can reach its goal period.
%
Yet, as expected, allocating more non-cooperative workgroups increases
the execution time overhead of the cooperative kernel.

@@

\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_ecology.pdf} \\
%\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_color_G3_circuit.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_mis_ecology.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_mis_G3_circuit.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_sssp_usa_road.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_sssp_rmat20.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_sssp_usa_ny.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bc_1k_128k.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bc_2k_1M.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bfs_rmat22.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd5500_bfs_usa.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_color_ecology.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_color_G3_circuit.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_mis_ecology.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_mis_G3_circuit.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_sssp_usa_road.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_sssp_rmat20.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_sssp_usa_ny.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_bc_1k_128k.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_bc_2k_1M.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_bfs_rmat22.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/iris_bfs_usa.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_color_ecology.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_color_G3_circuit.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_mis_ecology.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_mis_G3_circuit.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_sssp_usa_road.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_sssp_rmat20.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_sssp_usa_ny.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bc_1k_128k.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bc_2k_1M.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bfs_rmat22.pdf} \\
\includegraphics[width=.7\columnwidth]{images/barrier/hd520_bfs_usa.pdf} \\
\includegraphics[width=.7\columnwidth]{images/ws/hd5500_octree_NA.pdf} \\
\includegraphics[width=.7\columnwidth]{images/ws/hd5500_connect_four_NA.pdf} \\
\includegraphics[width=.7\columnwidth]{images/ws/iris_octree_NA.pdf} \\
\includegraphics[width=.7\columnwidth]{images/ws/iris_connect_four_NA.pdf} \\
\includegraphics[width=.7\columnwidth]{images/ws/hd520_octree_NA.pdf} \\
